\chapter{SYSTEM ANALYSIS}
 
\section{Module Description}
The Vehicle Bidding System is a comprehensive full-stack web application composed of multiple functional modules, each serving a critical role in the system's operation and user experience. The main modules include:

\begin{itemize}
    \item \textbf{Authentication and Authorization Module:} This module handles user registration, login, and session management. It implements JWT-based authentication to ensure secure communication between frontend and backend. Passwords are hashed using bcrypt for enhanced security. The module also manages role-based access control (RBAC), ensuring that users can only access features appropriate to their role (buyer, seller, or administrator).
    
    \item \textbf{Vehicle Management Module:} This module enables sellers to create, edit, and delete vehicle listings with all relevant specifications (brand, model, year, condition), pricing information, and document references. The module supports multiple image uploads and tracks vehicle status throughout the lifecycle (Available, Hidden, Sold).
    
    \item \textbf{Search and Browsing Module:} This module provides comprehensive search and filtering capabilities allowing buyers to discover vehicles based on multiple criteria including brand, model, year range, price range, vehicle category, fuel type, and transmission type.
    
    \item \textbf{Bidding System Module:} This is the core module that implements the real-time auction mechanism. It manages bid placement with validation rules, tracks the highest current bid, and maintains complete bid history with timestamps. The module uses Socket.io for real-time bid notifications and prevents bids lower than the current highest bid.
    
    \item \textbf{Messaging System Module:} This module facilitates real-time communication between buyers and sellers using Socket.io for instant message delivery. It stores message history and implements read status tracking.
    
    \item \textbf{Admin Dashboard Module:} This module provides administrators with system-wide visibility including user management, vehicle verification, user blacklisting capabilities, and comprehensive system analytics.
    
    \item \textbf{User Management Module:} This module handles all user-related operations including profile management, personal information updates, account settings, and user activity logs for security and compliance purposes.
\end{itemize}

\section{Business Rules}
Business rules are the governing principles and constraints that ensure the system operates according to business objectives and maintains data integrity. The key business rules include:

\subsection{User Management Rules}
\begin{itemize}
    \item All users must complete registration with unique email addresses
    \item Email addresses must be validated before account activation
    \item Passwords must meet security requirements
    \item Sellers must be verified before listing vehicles
    \item User accounts can be blacklisted by administrators for policy violations
\end{itemize}

\subsection{Vehicle Listing Rules}
\begin{itemize}
    \item All vehicle listings must be verified by administrators before publication
    \item Sellers can only edit or delete listings that have not received bids
    \item Each vehicle listing must include required specifications and at least one image
    \item Vehicle documents (RC, Insurance) must be properly documented
    \item Sold vehicles cannot be re-listed without admin approval
\end{itemize}

\subsection{Bidding Rules}
\begin{itemize}
    \item Bid amounts must be higher than the current highest bid
    \item Sellers cannot bid on their own vehicles
    \item Each bid is timestamped for tracking and dispute resolution
    \item Winners are determined when auction time expires
    \item Bid history is maintained for audit purposes
\end{itemize}

\subsection{Communication Rules}
\begin{itemize}
    \item Messages can only be exchanged between registered users
    \item Message history is preserved for dispute resolution
\end{itemize}

\subsection{Admin Rules}
\begin{itemize}
    \item Only administrators can verify vehicle listings
    \item Admin actions are logged for accountability
    \item Administrators can access system reports and analytics
\end{itemize}

\newpage
\section{System Architecture}

The Vehicle Bidding System follows a three-tier architecture pattern:

\subsection{Presentation Layer (Frontend)}
The React.js frontend provides a responsive user interface that works seamlessly across desktop and mobile devices. It handles all user interactions, form validation, and real-time updates through Socket.io. The frontend maintains component-based architecture for reusability and maintainability.

\subsection{Application Layer (Backend)}
The Express.js REST API server processes client requests, enforces business logic, manages authentication, and coordinates with the database. Socket.io enables real-time communication for bidding and messaging features. The backend implements middleware for authentication, error handling, and logging.

\subsection{Data Layer (Database)}
MongoDB provides flexible document storage appropriate for the system's data structures. Mongoose ODM provides schema validation and query building. The database is optimized with appropriate indexing on frequently queried fields (user IDs, vehicle IDs, bid timestamps).

\section{Technology Stack Justification}

\subsection{MongoDB}
Chosen for its flexibility in handling vehicle and bid data with varying structures, and its excellent performance for real-time queries required in bidding operations.

\subsection{Express.js and Node.js}
Provides a lightweight, efficient server capable of handling high concurrent connections required for real-time bidding, along with an extensive ecosystem of packages for authentication, socket programming, and database interaction.

\subsection{React.js}
Offers component reusability, efficient DOM manipulation through virtual DOM, and excellent tooling for building responsive user interfaces.

\subsection{Socket.io}
Enables real-time, bidirectional communication between client and server, essential for instant bid notifications and live messaging features.

\section{Project Pipeline}
\begin{figure}[H]
    \centering
    \begin{tikzpicture}[>=stealth, node distance=2.5cm, on grid]
        \node[draw, rectangle, minimum width=2.2cm] (plan) {Planning};
        \node[draw, rectangle, right=of plan] (design) {Design};
        \node[draw, rectangle, right=of design] (dev) {Development};
        \node[draw, rectangle, right=of dev] (test) {Testing};
        \node[draw, rectangle, below=of plan] (deploy) {Deployment};
        
        \draw[->] (plan) -- (design);
        \draw[->] (design) -- (dev);
        \draw[->] (dev) -- (test);
        \draw[->] (test) -- (deploy);
    \end{tikzpicture}
    \caption{Project Development Pipeline}
    \label{fig:pipeline}
\end{figure}

\section{Feasibility Analysis}

\subsection{Technical Feasibility}
The technology stack is well-established and thoroughly tested in production environments. All required libraries and frameworks are open-source with active communities. The development team has experience with these technologies, making technical implementation feasible within the planned timeline. The MERN stack is industry-standard for web applications requiring real-time communication.

\subsection{Economic Feasibility}
The use of open-source technologies significantly reduces licensing costs. Cloud hosting options like AWS, Heroku, or MongoDB Atlas provide cost-effective scalability. Development resources are available and cost-effective, making the project economically viable. The modular architecture allows for phased development, spreading costs across the project lifecycle.

\subsection{Operational Feasibility}
The system is designed to be user-friendly across all user types. Training requirements are minimal due to intuitive interface design. Administrative tools are provided for easy system management and monitoring. The REST API design allows for easy integration with future external services if needed.

\subsection{Schedule Feasibility}
With a well-defined scope, established technology stack, and experienced development team, the project can be completed within the planned timeline. The phased development approach allows for flexibility and risk mitigation. Approximately 4-6 months is required for complete development including testing and deployment.

\section{System Environment}

\subsection{Software Environment}

\textbf{Node.js and Express.js:}
Node.js provides the runtime environment for executing JavaScript on the server side. Express.js is a lightweight web framework built on Node.js that simplifies the development of REST APIs and middleware implementation.

\textbf{React.js:}
React.js is used to build the frontend Single Page Application (SPA) with component-based architecture. It provides efficient rendering through the virtual DOM and excellent tooling for state management and routing.

\textbf{MongoDB:}
MongoDB is a NoSQL database that stores application data in flexible JSON-like documents. It's chosen for its ability to handle varying data structures and excellent performance for real-time operations.

\textbf{Socket.io:}
Socket.io enables real-time, bidirectional communication between client and server using WebSockets, essential for live bidding notifications and instant messaging.

\textbf{Mongoose:}
Mongoose is an Object Data Modeling (ODM) library that provides schema validation and query building for MongoDB.

\textbf{JWT (JSON Web Tokens):}
JWT is used for secure user authentication and authorization without requiring server-side session storage.

\textbf{Bcrypt:}
Bcrypt is used for secure password hashing to protect user credentials.

\textbf{Vite:}
Vite is a modern build tool that provides fast development server and optimized production builds for the React frontend.

\subsection{Hardware Environment}

\textbf{Development Machine:}
\begin{itemize}
    \item Processor: Intel i5 12th Gen or equivalent
    \item RAM: 16 GB or more
    \item Storage: 512 GB SSD
    \item Graphics: Integrated graphics sufficient for web development
\end{itemize}

\textbf{Server Environment (Production):}
\begin{itemize}
    \item Cloud hosting: AWS EC2, Heroku, or similar
    \item Load balancing for handling concurrent users
    \item Database: MongoDB Atlas or self-hosted MongoDB
    \item CDN for static asset delivery
\end{itemize}

\subsection{Browser Compatibility}
\begin{itemize}
    \item Chrome (latest versions)
    \item Firefox (latest versions)
    \item Safari (latest versions)
    \item Edge (latest versions)
\end{itemize}

\section{Actors and Roles}

\subsection{Buyer}
\textbf{Role:} Browse vehicles, place bids, communicate with sellers, track bidding history
\begin{itemize}
    \item Can search and filter vehicles
    \item Can place bids on available vehicles
    \item Can message sellers for inquiries
    \item Can view their bid history
    \item Can track vehicles they're interested in
\end{itemize}

\subsection{Seller}
\textbf{Role:} List vehicles, manage listings, respond to inquiries, track bids
\begin{itemize}
    \item Can create vehicle listings with specifications
    \item Can upload multiple images and documents
    \item Can edit listings before bidding starts
    \item Can view and respond to buyer inquiries
    \item Can monitor bids on their vehicles
\end{itemize}

\subsection{Administrator}
\textbf{Role:} Verify listings, manage users, generate reports, ensure system integrity
\begin{itemize}
    \item Can verify or reject vehicle listings
    \item Can manage user accounts and permissions
    \item Can blacklist users for policy violations
    \item Can access system statistics and reports
    \item Can view system logs and monitor platform activity
\end{itemize}

\subsection{System (Automated)}
\textbf{Role:} Execute auction logic, send notifications, validate data
\begin{itemize}
    \item Determines winning bids when auction ends
    \item Sends real-time bid notifications
    \item Validates bid amounts and user permissions
    \item Maintains data consistency and backups
\end{itemize}

